Experience shapes individual foraging specialization and success in a virtual predator-prey system

### 2023-10-26 manuscript/manuscript.Rmd
minor updates to methods and last par in intro

Predator express a diverse array of hunting tactics to capture their prey. In natural populations, differences among individuals in tactics such as ambushing, fast movement speeds, or high attack frequencies are associated with differences in prey consumption [@toscano2014;@patrick.weimerskirch2014a;@toscano.etal2016]. Prey usually respond to predation cues by reducing their activity, or use rapid or unpredictable movements to successfully escape when they are chased [@walker.etal2005;@kelley.magurran2011;@herbert-read.etal2017]. This results in reciprocal interactions where predators fine-tune their tactics to their prey, suggesting that they learn and become efficient hunters through repeated encounters [@estes.etal2003;@woo.etal2008;@phillips.etal2017]. These behavioural adjustments with experience can shape the outcome of predator-prey interactions, which may result in larger implications for community and ecosystem dynamics [@wooster.etal2023].

@@

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Individual-based models can mitigate some of these challenges and have contributed in stirring testable ideas for empiricists [@cornell2019]. Yet, while providing maximum control over environmental parameters, they can suffer from emerging behaviours that are too complex to analyze mathematically [@cornell2019]. Videogames offer an interesting alternative to agent-based simulations, providing controled virtual environments where real players interact. For example, in the predator-prey videogame *Dead by Daylight*, four prey players need to forage for resources while avoiding predation by a fifth player. In this virtual system, the predator population is composed of individuals that either ambush or hunt at high speeds (i.e. mean movement speed along a slow-fast continuum), and their success is driven by the movement of the prey [@fraserfranco.etal2022]. We found that prey can increase their chances of survival by cooperating and moving fast when they forage [@cere.etal2021;@, Santostefano et al. in press]. Of course, the life of neither the predator nor the prey players are at stake, such that emerging patterns could be driven by their motivation to win and not "true" survival. Yet, the game ellicits natural reactions in players such as fear from predation (personal observations), which corroborates with another virtual ecological study showing that predation drives individual variation in risk perception [@beauchamp2020]. In a different system, @lymbery.etal2023 showed that complex environments favor strong soldiers over large armies in a virtual game as well as in the laboratory with two ant species. These observations outline how ecological patterns can emerge from interactions in virtual systems with fixed rules. Videogames also generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, virtual systems such as *Dead by Daylight* offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

In this study, we assess how repeated experience shapes predator hunting specialization and success using data from players in *Dead by Daylight*. First, we investigate how individual hunting success changes with experience (i.e. how individuals develop their hunting expertise). We hypothesize that the predators' success will increase with experience up to a certain level where it will stagnate. We expect this pattern to vary among individuals due to differences in prey encounters. For example, a predator may face greater difficulty than other individuals if it encountered elusive groups of prey more often. Second, we assess how foraging specialization emerges with experience. We define the level of individual foraging specialization as the intra-individual variance (i.e. IIV) in movement speed across matches, where low IIV translates to individual specialization (i.e. always using a cursorial tactic). If all predators encounter varying groups of prey, we predict that individuals (and thus the population) should all converge towards flexible speeds. However, if individuals encounter similar groups of prey, then everyone should instead specialize in similar speeds. In both scenarios, among individual variation in IIV (i.e. individual differences in specialization) should be low across experience, whereas the population variance would either increase (flexibility) or decrease (specialization). Alternatively, individual specialization and flexibility may emerge if individuals experience different interactions with their prey, such that among individual variation in IIV would increase. In this case, we expect predators that experienced more predictable encounters with their prey (i.e. similar groups across matches) to specialize in similar speeds, while predators that experienced unpredictable encounters with their prey (i.e. heterogeneous groups across matches) should adopt flexible movement speeds. If we detect such prey-dependent fine-tuning, then all hunters along the flexible-specialist continuum should attain equal success if they learned from experience how to adjust their hunting style to their prey.

@@

We analyzed the predator's mean speed and the mean speed of the prey group encountered by the predator. The predator's mean speed is measured as the mean distance traveled per second during a match (mean= `r toString(format(round(mean(dat$pred_speed), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$pred_speed), digits = 2), nsmall = 2))` m/s). We measured the preys' speed as the mean travel speed of the four individual prey in a match (mean = `r toString(format(round(mean(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` m/s). We defined hunting success as the number of prey consumed during the match (min = `r toString(format(min(dat$hunting_success)))`, max = `r toString(format(max(dat$hunting_success)))`). Lastly, we defined the predator's cumulative experience as the number of matches played prior to the match being monitored. For example, the first match of a player would have a cumulative experience value of 0, while the tenth match would have a value of 9. We also categorized predators for each match based on their cumulated experience. We labeled predators as novices for matches where they had cumulated less than 100 matches, intermediate for matches where they had cumulated between 100 and 299 matches, and advanced for matches where they had cumulated more than 299 matches (max 499). Since our goal was to monitor predator players throughout their experience and that they all played at least 300 matches, they all appeared in each of the three experience categories.

@@

The predator population maintained its speed and degree of specialization throughout the study period, and individuals differed only slightly in their average speeds. However, differences in specialization among individuals were important and increased slightly with experience. Predators hunting at higher speeds specialized in this tactic and tended to encounter groups of prey with similar speeds, while those hunting at slower speeds were flexible and tended to encounter varying groups of prey. On one hand, this pattern could arise if some predators adjusted their strategy to their prey, because fast-paced hunting is a specialized tactic suited for prey that use rapid evasive movements, while a slower and flexible tactic can be useful to minimize the consequences of uncertainty when prey escape unpredictably [@endler1991;@bro-jorgensen2013;@wilson.etal2018;@szopa-comley.ioannou2022]. However, we observed that 44% of the predator population showed close to no change in specialization with experience. Since these individuals were already specialized as novices (Fraser Franco, personal observation), there would be no point in them changing their tactic with experience if they were already successful, or there could have been costs to switch their strategy if they more often encountered faster prey. This could explain why the behavioural interactions between the predator and the prey remained stable across the predator's experience. Yet, because prey can also learn how to avoid predation [@kelley2003;@turner2006;@mccomb2011], we believe that the predator-prey phenotype matching more likely emerged from reciprocal adjustments in speed by predators and prey as they interacted [@kishida.etal2006;@kishida.etal2009;@edgell.rochette2009;@mcghee.etal2013]. Indeed, 56% of the predator population altered its behaviour with experience to different degrees, including 12% that even switched between flexibility and specialization. Hence, if the prey also learned through repeated interactions with the predators, it is possible that experience contributed in stabilizing the system as both were adjusting to each other, similar to Red Queen dynamics [@brockhurst2014]. Our results appear to support this since many individuals were shifting their behavioural niche while the population niche remained stable across experience.

@@

A general assumption of predator-prey studies is that predators maximize success by matching their tactic to their prey [@abrams2000], yet, it has remained largely unknown whether this results from predators learning how to hunt their prey, in part because of the challenges of investigating direct interactions in the wild. By capitalizing on a virtual predator-prey system where interactions were directly monitored, we found support of our prediction that prey were driving individual differences in expertise in a predator population. We also confirmed that experience and predator-prey encounters drove the patterns of individual foraging specialization and flexibility. Our results suggest that predators learned with experience, as their success increased and their speed remained matched to the speed of their prey. Even though individuals were not all equally successful, both specialized and flexible hunters achieved similar success. One limitation of our study was that we couldn't monitor all the matches of the prey, which prevented us from assessing their responses to the predator. Hence, future studies should aim at monitoring reciprocal behavioural dynamics through time, which may reveal important insights on the mechanisms driving stable equilibria in predator-prey systems. Lastly, virtual systems are increasingly recognized among ecologists as being useful to test hypotheses on consumer-resource interactions [@beauchamp2020;@cere.etal2021;@fraserfranco.etal2022;@lymbery.etal2023]. We therefore hope that our study will inspire more collaborations between scientists and the videogame industry to tackle fundamental questions in ecology.

>>>>

Predator express a diverse array of hunting tactics to capture their prey. In natural populations, differences among individuals in tactics such as ambushing and fast movement speeds are associated with differences in prey consumption [@toscano2014;@patrick.weimerskirch2014a;@toscano.etal2016]. Prey usually respond to predation cues by reducing their activity, or use rapid or unpredictable movements to successfully escape when they are chased [@walker.etal2005;@kelley.magurran2011;@herbert-read.etal2017]. This results in reciprocal interactions where predators fine-tune their tactics to their prey, suggesting that they learn and become efficient hunters through repeated encounters [@estes.etal2003;@woo.etal2008;@phillips.etal2017]. These behavioural adjustments with experience can shape the outcome of predator-prey interactions, which may result in larger implications for community and ecosystem dynamics [@wooster.etal2023].

@@

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Individual-based models can mitigate some of these challenges and have contributed in stirring testable ideas for empiricists [@cornell2019]. Yet, while providing maximum control over environmental parameters, they can suffer from emerging behaviours that are too complex to analyze mathematically [@cornell2019]. Videogames offer an interesting alternative to agent-based simulations, providing controled virtual environments where real players interact. For example, in the predator-prey videogame *Dead by Daylight*, four prey players need to forage for resources while avoiding predation by a fifth player. In this virtual system, the predator population is composed of individuals that either ambush or hunt at high speeds (i.e. mean movement speed along a slow-fast continuum), and their success is driven by the movement of the prey [@fraserfranco.etal2022]. We found that prey can increase their chances of survival by cooperating and moving fast when they forage [@cere.etal2021;@fraserfranco.etal2022, Santostefano et al. in press]. Of course, the life of neither the predator nor the prey players are at stake, such that emerging patterns could be driven by their motivation to win and not "true" survival. Yet, the game ellicits natural reactions in players such as fear from predation (personal observations), which corroborates with another virtual ecological study showing that predation drives individual variation in risk perception [@beauchamp2020]. In a different system, @lymbery.etal2023 showed that complex environments favor strong soldiers over large armies in a virtual game as well as in the laboratory with two ant species. These observations outline how ecological patterns can emerge from interactions in virtual systems with fixed rules. Videogames also generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, virtual systems such as *Dead by Daylight* offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

In this study, we assess how repeated experience shapes predator hunting specialization and success using data from players in *Dead by Daylight*. Following @fraserfranco.etal2022 we use the predators' mean movement speed along a slow-fast continuum as a proxy for the hunting tactic (ambush-cursorial), and define the level of individual hunting specialization as the intra-individual variance (i.e. IIV) in movement speed across matches, where low IIV translates to individual specialization (i.e. always using a cursorial tactic). First, we investigate how individual hunting success changes with experience (i.e. how individuals develop their hunting expertise). We hypothesize that the predators' success will increase with experience up to a certain level where it will stagnate. We expect this pattern to vary among individuals due to differences in prey encounters. Second, we assess how foraging specialization emerges with experience. If all predators encounter varying groups of prey, we predict that individuals (and thus the population) should all converge towards flexible speeds. However, if individuals encounter similar groups of prey, then everyone should instead specialize in similar speeds. In both scenarios, among individual variation in IIV (i.e. individual differences in specialization) should be low across experience, whereas the population variance would either increase (flexibility) or decrease (specialization). Alternatively, individual specialization and flexibility may emerge if individuals experience different interactions with their prey, such that among individual variation in IIV would increase. In this case, we expect predators that experienced more predictable encounters with their prey (i.e. similar groups across matches) to specialize in similar speeds, while predators that experienced unpredictable encounters with their prey (i.e. heterogeneous groups across matches) should adopt flexible movement speeds. If we detect such prey-dependent fine-tuning, then all hunters along the flexible-specialist continuum should attain equal success if they learned from experience how to adjust their hunting style to their prey.

@@

We analyzed the predator's mean speed and the mean speed of the prey group encountered by the predator. The predator's mean speed is measured as the mean distance traveled per second during a match (mean= `r toString(format(round(mean(dat$pred_speed), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$pred_speed), digits = 2), nsmall = 2))` m/s). We measured the preys' speed as the mean travel speed of the four individual prey in a match (mean = `r toString(format(round(mean(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` m/s). We defined hunting success as the number of prey consumed during the match (min = `r toString(format(min(dat$hunting_success)))`, max = `r toString(format(max(dat$hunting_success)))`). Lastly, we defined the predator's cumulative experience as the number of matches played as the predator prior to the match being monitored. For example, the first match of a player would have a cumulative experience value of 0, while the tenth match would have a value of 9. We did not account for matches where predators played as the prey. We also categorized predators for each match based on their cumulated experience. We labeled predators as novices for matches where they had cumulated less than 100 matches, intermediate for matches where they had cumulated between 100 and 299 matches, and advanced for matches where they had cumulated more than 299 matches (max 499). Since our goal was to monitor predator players throughout their experience and that they all played at least 300 matches, they all appeared in each of the three experience categories.

@@

The predator population maintained its speed and degree of specialization throughout the study period, and individuals differed only slightly in their average speeds. However, differences in specialization among individuals were important and increased slightly with experience. Predators hunting at higher speeds specialized in this tactic and tended to encounter groups of prey with similar speeds, while those hunting at slower speeds were flexible and tended to encounter varying groups of prey. On one hand, this pattern could arise if some predators adjusted their strategy to their prey, because fast-paced hunting is a specialized tactic suited for prey that use rapid evasive movements, while a slower and flexible tactic can be useful to minimize the consequences of uncertainty when prey escape unpredictably [@endler1991;@bro-jorgensen2013;@wilson.etal2018;@szopa-comley.ioannou2022]. However, we observed that 44% of the predator population showed close to no change in specialization with experience. Since these individuals were already specialized as novices (Fraser Franco, personal observation), there would be no point in them changing their tactic with experience if they were already successful, or there could have been costs to switch their strategy if they more often encountered faster prey. This could explain why the behavioural interactions between the predator and the prey remained stable across the predator's experience. Yet, because prey can also learn how to avoid predation [@kelley2003;@turner2006;@mccomb2011], we believe that the predator-prey phenotype matching more likely emerged from reciprocal adjustments in speed by predators and prey as they interacted [@kishida.etal2006;@kishida.etal2009;@edgell.rochette2009;@mcghee.etal2013]. Indeed, 56% of the predator population altered its behaviour with experience to different degrees, including 12% that even switched between flexibility and specialization. Hence, if the prey also learned through repeated interactions with the predators, it is possible that experience contributed in stabilizing the system as both were adjusting to each other, similar to Red Queen dynamics [@brockhurst2014].

@@

A general assumption of predator-prey studies is that predators maximize success by matching their tactic to their prey [@abrams2000], yet, it has remained largely unknown whether this results from predators learning how to hunt their prey in part because of the challenges of investigating direct interactions in the wild. By capitalizing on a virtual predator-prey system where interactions were directly monitored, we found support of our prediction that prey were driving individual differences in expertise in a predator population. We also confirmed that experience and predator-prey encounters drove the patterns of individual foraging specialization and flexibility. Our results suggest that predators learned with experience, as their success increased and their speed remained matched to the speed of their prey. Even though individuals were not all equally successful, both specialized and flexible hunters achieved similar success. One limitation of our study was that we couldn't monitor all the matches of the prey, which prevented us from assessing their responses to the predator. Hence, future studies should aim at monitoring reciprocal behavioural dynamics through time, which may reveal important insights on the mechanisms driving stable equilibria in predator-prey systems. Lastly, virtual systems are increasingly recognized among ecologists as being useful to test hypotheses on consumer-resource interactions [@beauchamp2020;@cere.etal2021;@fraserfranco.etal2022;@lymbery.etal2023]. We therefore hope that our study will inspire more collaborations between scientists and the videogame industry to tackle fundamental questions in ecology.
